Supporting Information:\\Black-swan events in animal populations

### 2016-07-13 text/ms_pnas.tex
Cleanup
Switch panel letters in all captions to uppercase

\begin{figure*}[htb]
\centering
\includegraphics[width=0.65\textwidth]{../analysis/t-nu-eg2.pdf}
\caption{ Illustration of population dynamic models that allow for heavy
tails. (a, b) Probability density for the Student-$t$
distribution with scale parameter of $1$ and different values of \(\nu\). Small
values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
distribution approaches the normal distribution. (c to e)
Simulated population dynamics from a Gompertz model with process noise drawn
from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
panels c and d represent jumps with less than a 1 in 1000
chance of occurring in a normal distribution. (f to h)
Estimates of \(\nu\) from models fit to the times series in panels c to
e. Shown are posterior samples (histograms), median and interquartile
range of the posterior (IQR, dots and line segments), and the exponential prior
on \(\nu\) (dashed lines). Colour shading behind panels f to h
illustrates the approximate region of heavy tails.}
\label{fig:1}
\end{figure*}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.3\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{Estimates of population dynamic heavy-tailedness for
populations of birds, mammals, insects, and fishes. (a to d)
Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
points and line segments represent posterior medians and 50\% / 90\% credible
intervals for individual populations. Inset plots show probability that \(\nu
< 10\) for populations arranged by taxonomic order and sorted by decreasing
mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
panel a are omitted for space. Red to yellow points highlight populations
with a high to moderately high probability of heavy-tailed dynamics.
}
\label{fig:2}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.26\textwidth]{../analysis/order-posteriors-covariates.pdf}
\caption{Standardized probabilities and covariates of heavy-tailed
dynamics. (a) Taxonomic-order-level posterior densities of Pr(\(\nu
< 10\)) after accounting for time-series length in a hierarchical model. Dotted
vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the
prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
green: insects, purple: birds, and red: mammals). Estimates are standardized at
the geometric mean of time series length across all the data (approximately 27
time steps) and shown for orders with \(\ge 5\) populations. (b)
Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
panels, short vertical line segments within the density polygons indicate
medians.
}
\label{fig:3}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.45\textwidth]{../analysis/skew-fig.pdf}
\caption{Heavy-tailed process noise tends to be downward skewed and
ignoring this can underestimate risk. (a) Illustrations of Student-$t$
distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
(b) Posterior density of the skewness parameters aggregated across
populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly
heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations
(\(\hat{\nu} > 70\)). Approximate mid-values from b are illustrated in
panel a. (c) Example time series of heavy-tailed populations with
different levels of skewness. Red dots highlight likely heavy-tailed events.
Forecasts (grey regions) show median (solid lines) and lower 99\% credible
intervals (dotted lines) of abundance. Black and red lines indicate forecasts
from Gompertz models with lognormal and skew-$t$ process noise, respectively.
}
\label{fig:4}
\end{figure}

@@

% \section*{Figures}

% \clearpage

% \clearpage

% \clearpage

>>>>

\begin{figure*}[htb]
\centering
\includegraphics[width=0.65\textwidth]{../analysis/t-nu-eg2.pdf}
\caption{ Illustration of population dynamic models that allow for heavy
tails. (A, B) Probability density for the Student-$t$
distribution with scale parameter of $1$ and different values of \(\nu\). Small
values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
distribution approaches the normal distribution. (C to E)
Simulated population dynamics from a Gompertz model with process noise drawn
from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
panels C and D represent jumps with less than a 1 in 1000
chance of occurring in a normal distribution. (F to H)
Estimates of \(\nu\) from models fit to the times series in panels C to
E. Shown are posterior samples (histograms), median and interquartile
range of the posterior (IQR, dots and line segments), and the exponential prior
on \(\nu\) (dashed lines). Colour shading behind panels F to H
illustrates the approximate region of heavy tails.}
\label{fig:1}
\end{figure*}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.3\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{Estimates of population dynamic heavy-tailedness for
populations of birds, mammals, insects, and fishes. (A to D)
Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
points and line segments represent posterior medians and 50\% / 90\% credible
intervals for individual populations. Inset plots show probability that \(\nu
< 10\) for populations arranged by taxonomic order and sorted by decreasing
mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
panel a are omitted for space. Red to yellow points highlight populations
with a high to moderately high probability of heavy-tailed dynamics.
}
\label{fig:2}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.26\textwidth]{../analysis/order-posteriors-covariates.pdf}
\caption{Standardized probabilities and covariates of heavy-tailed
dynamics. (A) Taxonomic-order-level posterior densities of Pr(\(\nu
< 10\)) after accounting for time-series length in a hierarchical model. Dotted
vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the
prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
green: insects, purple: birds, and red: mammals). Estimates are standardized at
the geometric mean of time series length across all the data (approximately 27
time steps) and shown for orders with \(\ge 5\) populations. (B)
Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
panels, short vertical line segments within the density polygons indicate
medians.
}
\label{fig:3}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.45\textwidth]{../analysis/skew-fig.pdf}
\caption{Heavy-tailed process noise tends to be downward skewed and
ignoring this can underestimate risk. (A) Illustrations of Student-$t$
distributions with three levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
(B) Posterior density of the skewness parameters aggregated across
populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly
heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations
(\(\hat{\nu} > 70\)). Approximate mid-values from B are illustrated in
panel A. (C) Example time series of heavy-tailed populations with
different levels of skewness. Red dots highlight likely heavy-tailed events.
Forecasts (grey regions) show median (solid lines) and lower 99\% credible
intervals (dotted lines) of abundance. Black and red lines indicate forecasts
from Gompertz models with lognormal and skew-$t$ process noise, respectively.
}
\label{fig:4}
\end{figure}

@@

### 2016-07-13 text/supp.tex
Switch panel letters in all captions to uppercase

\caption{Probability density of the Bayesian priors for the Gompertz models. \textbf{(a)} Per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$. \textbf{(b)} Scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ \textbf{(c)} Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$. \textbf{(d)} AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$. Panel \textbf{(c)} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$. The inset panel shows the same data but with a log-transformed x axis. Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ with which we are concerned.}

@@

\caption{Estimates of $\nu$ from alternative models plotted against the
    base Gompertz model estimates of $\nu$. Alternative models are \textbf{(a)}
    Ricker-logistic; \textbf{(b)} a Gompertz model with AR1 residual correlation;
    \textbf{(c)} modelling population growth rate with no density dependence; and
    \textbf{(d)} a Gompertz model with fixed magnitude of observation error. Shown
are medians of the posterior (dots) and 50\% credible intervals (segments). The
diagonal line indicates a one-to-one relationship. Different colours indicate
various taxonomic classes. The grey-shaded regions indicate regions of
disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
The Gompertz observation error model assumes a fixed standard deviation of
observation error of $0.2$ on a log scale. Models in \textbf{(b)} and \textbf{(d)}
that had not met convergence thresholds were run for four chains, 256,000 iterations (128,000 warm up), and retaining every 10th sample, after which the MCMC chains for one population in \textbf{(b)} and four populations in \textbf{(d)} had still not met convergence thresholds. These populations are excluded from their respective panels, but all populations with Pr($\nu < 10$) $> 0.5$ in the base model had converged and are shown.}

@@

\caption{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$. \textbf{(a)} A weaker prior on $\nu$; \textbf{(b)} a stronger prior on $\nu$, as illustrated in Fig.~1. Shown are medians of the posterior (dots) and 50\% credible intervals (segments). The diagonal line indicates a one-to-one relationship. Different colours indicate various taxonomic classes. The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics. In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$. When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.}

@@

\caption{Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$). Shown are (\textbf{a} to \textbf{c}) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), \textbf{(d)} number of time steps, \textbf{(e)} body length, and \textbf{(f)} lifespan. For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-$t$ process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter. Circles representing a few sharks, crustaceans, and gastropods are filled in white. Median and 90\% credible interval posterior predictions of a beta regression hierarchical model are shown in panels \textbf{(a)} and \textbf{(d)} where there was a high probability the slope coefficient was different from zero (Fig.~3B).}

>>>>

\caption{Probability density of the Bayesian priors for the Gompertz models. \textbf{(A)} Per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$. \textbf{(B)} Scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ \textbf{(C)} Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$. \textbf{(D)} AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$. Panel \textbf{(C)} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$. The inset panel shows the same data but with a log-transformed x axis. Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ with which we are concerned.}

@@

\caption{Estimates of $\nu$ from alternative models plotted against the
    base Gompertz model estimates of $\nu$. Alternative models are \textbf{(A)}
    Ricker-logistic; \textbf{(B)} a Gompertz model with AR1 residual correlation;
    \textbf{(C)} modelling population growth rate with no density dependence; and
    \textbf{(D)} a Gompertz model with fixed magnitude of observation error. Shown
are medians of the posterior (dots) and 50\% credible intervals (segments). The
diagonal line indicates a one-to-one relationship. Different colours indicate
various taxonomic classes. The grey-shaded regions indicate regions of
disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
The Gompertz observation error model assumes a fixed standard deviation of
observation error of $0.2$ on a log scale. Models in \textbf{(B)} and \textbf{(D)}
that had not met convergence thresholds were run for four chains, 256,000 iterations (128,000 warm up), and retaining every 10th sample, after which the MCMC chains for one population in \textbf{(B)} and four populations in \textbf{(D)} had still not met convergence thresholds. These populations are excluded from their respective panels, but all populations with Pr($\nu < 10$) $> 0.5$ in the base model had converged and are shown.}

@@

\caption{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$. \textbf{(A)} A weaker prior on $\nu$; \textbf{(B)} a stronger prior on $\nu$, as illustrated in Fig.~1. Shown are medians of the posterior (dots) and 50\% credible intervals (segments). The diagonal line indicates a one-to-one relationship. Different colours indicate various taxonomic classes. The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics. In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$. When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.}

@@

\caption{Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$). Shown are (\textbf{A} to \textbf{C}) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), \textbf{(D)} number of time steps, \textbf{(E)} body length, and \textbf{(F)} lifespan. For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-$t$ process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter. Circles representing a few sharks, crustaceans, and gastropods are filled in white. Median and 90\% credible interval posterior predictions of a beta regression hierarchical model are shown in panels \textbf{(A)} and \textbf{(D)} where there was a high probability the slope coefficient was different from zero (Fig.~3B).}
